Information carried by thalamocortical axons enables self-organization of the barrel field

### 2019-11-28 paper/briefpaper/brief.tex
Now have a _medium_ length paper, mostly finished

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.
%(\citealp{anton-bolanos_prenatal_2019}, see also
%\citealp{anton-bolanos_developmental_2018}).
Patterning also depends on a complex network of axon guidance processes, which
includes interactions between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{bishop_regulation_2000}, and the
binding of p75 to neurotropins
\citep{bishop_distinct_2002}.
% or both citations?
%\citep{shimogori_fibroblast_2005,bishop_distinct_2002}.
%
%(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}
%; see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}).
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17
and transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate \cite{assimacopoulos_fibroblast_2012}.
% A longer list of citations:
%\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008}, a complex gene
expression pattern with the complexity of the barrel field has not, to our
knowledge been reported and so the mechanism by which the barrels form seems
likely to be one in which the structure forms by self-organization.

@@

The dynamics of the model (Eqs.~\ref{eq:dc} \& \ref{eq:da}) were solved on a
hexagonal lattice of grid elements representing the cortical sheet, under the
assumption that axon branches cannot extend beyond the boundary of the sheet,
i.e., $\mb{J}_i(\mb{x},t) \big\rvert_{boundary} = 0$. We developed software
tools for imposing arbitrary boundary shapes on the lattice. The dynamics were
numerically integrated using fourth-order Runge-Kutta, and using a
{\color{red} finite difference method} to compute the flux term (see
Supplementary Materials).

@@

@@

To investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model. There does exist an element
of competition in the original model (Eqs.~\ref{eq:dc} \& \ref{eq:da}). The
term $\left(1 - \sum_{i=1}^{N} c_i\right)$ in Eq.~\ref{eq:dc} ensures that
once the available synaptic receptors in a region have been fully occupied by
one thalamocortical axon type, there is no longer the possibility for other
types to form synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region; it
is possible for two or more TC axon types to have roughly equal branching
density in some region, leading to roughly equal numbers of connections
forming for each TC type, up until the `hard' connection limit is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} f(a_{i'}).
\end{equation}
%
with the choice $f(a) = a$, which leads to a reduction in the rate of increase
of branching density for axon type $i$ in regions where the branching density
of \emph{other} axon types is high. The new parameter $\epsilon$ governs the
strength of the branching competition and the factor $1/(N-1)$ ensures that the
strength scales with $N$.

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area. It is then multiplied by the branching density per unit area at
time 0, to account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:comp} on the
basis of the system state at time $t-{\delta}t$.

@@

In which we show that a realistic barrel pattern in the thalamus can be
recreated in barrel cortex.

\cmnt{The realistic barrel pattern experiment is enhanced by stronger
  branching competition:}

To increase the strength of the competition, we choose an alternative form for
$f(a)$ in Eq.~\ref{eq:comp} which has a geometric form for small $a$: $f(a) = \frac{1}{1 + e^{-m[a]^l}} - \frac{1}{2}$. The parameters $l$ and $m$ control the steepness and shape of the
function.

%%% End of stronger branching competition %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./MainFig.png}
\end{center}
%\vspace{-0em}
\caption{A. Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear. B. }
\label{fig:main}
\end{figure*}

@@

To cover the reaction to changes in the guidance molecules caused by
experiments like those by Shimogori \& Grove.

@@

We set out to test the hypothesis that barrel patterning may emerge via
self-organising processes. We first extended an existing model of neocortical
arealization to simulate pattern formation in two dimensions, inheriting from
this model the assumption that axonal branching and synaptogenesis are
coupled, and that thalamic projections compete for cortical territory under
the influence of molecular signalling gradients. Realistic barrel patterning
emerged only when terms were added to represent the assumption that
thalamocortical projections compete for a limited supply of connections with
cortical cells, that their branching rates are reduced by the branching of
other thalamocortical projections, and that the density of axon branching is
conserved for each projection. The net effect of these local interactions is a
Voronoi tesselation in which localised clusters of thalamic projections become
somatotopically organised on the cortical sheet according to the relative
values of coefficients specifying the rates at which axons are attracted and
repelled by orthogonal pairs of molecular signal gradients. According to this
model, the pattern of barrels is determined by the competitive and cooperative
interactions between cells relative to the shape of the cortical field
boundary. In simulation, we systematically investigated the origin of the
asymmetry between the barrel rows and arcs, and identified {\color{red} XXXXX}
as the most likely candidate for the elongation of barrel shapes along the
rows.

>>>>

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.
%(\citealp{anton-bolanos_prenatal_2019}, see also
%\citealp{anton-bolanos_developmental_2018}).
Patterning also depends on a complex network of axon guidance processes, which
includes interactions between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{bishop_regulation_2000}, and the
binding of p75 to neurotropins
\citep{bishop_distinct_2002}.
% or both citations?
%\citep{shimogori_fibroblast_2005,bishop_distinct_2002}.
%
%(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}
%; see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}).
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17
and transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate \cite{assimacopoulos_fibroblast_2012}.
% A longer list of citations:
%\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008}, we are not aware of
any report of gene expression patterns with the complexity of the barrel field
and so the mechanism by which the barrels form seems likely to be one in which
the structure forms by self-organization.

@@

The dynamics of the model (Eqs.~\ref{eq:dc} \& \ref{eq:da}) were solved on a
hexagonal lattice of grid elements representing the cortical sheet, under the
assumption that axon branches cannot extend beyond the boundary of the sheet,
i.e., $\mb{J}_i(\mb{x},t) \big\rvert_{boundary} = 0$. We developed software
tools for imposing arbitrary boundary shapes on the lattice. The dynamics were
numerically integrated using fourth-order Runge-Kutta (with fixed time-step
${\delta}t$), and using Gauss's Theorem to compute the flux term.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./MainFig.png}
\end{center}
%\vspace{-0em}
\caption{Simulated barrel fields. \textbf{A.} Barreloid centres taken from
  \cite{haidarliu_size_2001} are shown as black circles. Barreloids are
  labelled according to the usual convention. Labels R1 to R5 are the rhinal
  whisker barreloids. The dashed grey line represents the boundary of the
  barreloid field; the dot-dash line shows the anterior-posterior axis. It is
  assumed that there are two linearly increasing, orthogonal signalling
  gradients (molecules T1 \& T2) in the thalamus (side graphs), such that the
  location of each barreloid within the graded thalamic signal field can
  differentiate the efferent axons, with the result that axons from each
  barreloid interact to varying degrees with cortical signalling gradients
  (shown in panel B). Coloured arrows show, for each barreloid, the magnitude
  (and sign) of the interaction that its thalamocortical projection will have
  with the signalling gradients in the cortex. The mappings from barreloid
  location to interaction strength are linear. \textbf{B.} Guidance molecules
  in the simulated cortex. Molecule 1 is expressed high to low in an
  approximately anterior to posterior gradient. Molecule 2 is expressed
  lateral to medial (analogously to ephrin - Vanderhaegen et al) \textbf{C.}
  Whisker barrels from \cite{shimogori_fibroblast_2005}, with annotations for
  barrel identity and approximate axes.  \textbf{D-F.} Barrel cortex
  simulation. The maps indicate which thalomocortical axons make maximum
  connection density. Each hex element is coloured one of 52 unique colours
  depending on which e thalamocortical axon makes the most connections in that
  element. The mapping between barrel identity and colour is given in F.
  \textbf{G.} The Honda Voronoi measure as a function of simulation time for
  results shown in D--F. \textbf{H.}  Guidance molecules in the simulated Fgf8
  misexpression experiment. Molecule 1 is mirrored over the anterior-posterior
  axis \textbf{I.} The simulation indicates that the mirroring of the barrel
  field can occur as a result of the mirroring of guidance molecule 1 and the
  field boundary.}
\label{fig} % There's only one fig.
\end{figure*}

@@

To investigate how the whisker barrel structure (Fig.~\ref{fig}C) might
form in the presence of only monotonically varying morphogen gradients, we
turned our attention to the competition for connections built into the
model. There \emph{does} exist an element of competition in the original model
(Eqs.~\ref{eq:dc} \& \ref{eq:da}). The term $\left(1 - \sum_{i=1}^{N}
c_i\right)$ in Eq.~\ref{eq:dc} ensures that once the available synaptic
receptors in a region have been fully occupied by one thalamocortical axon
type, there is no longer the possibility for other types to form
synapses. This competition turns out to be relatively weak, because it does
not strongly affect the branching density in that region; it is possible for
two or more TC axon types to have roughly equal branching density in some
region, leading to roughly equal numbers of connections forming for each TC
type, up until the `hard' connection limit is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} a_{i'}
\end{equation}
%
%with the choice $f(a) = a$,
which leads to a reduction in
%the rate of increase of branching density for axon type $i$
%$\frac{\partial a_i}{\partial t}$
${\partial a_i}/{\partial t}$
in regions where the branching density
of \emph{other} axon types is high. The new parameter $\epsilon$ governs the
strength of the branching competition and the factor $1/(N-1)$ ensures that the
strength scales with $N$.

With the addition of branching competition, it is necessary to introduce a
normalisation step, in order to prevent one TC type from out-competing every
other type. This represents the idea that for a fixed number of afferent
thalamocortical axons, the overall number of branches must have some upper
limit, given the finite resources available in each cell for building new
processes. We divisively normalize the branching density, dividing it at each
computational step by the mean branching density per unit area. It is then
multiplied by the branching density per unit area at time 0, which preserves
the effect of any disparity in the numbers of axons of each type which
innervate the cortical sub-plate:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:comp} on the basis of
the system state at time $t-{\delta}t$. Integrals over $\Omega$ denote a
spatial integral over the domain of interest.

@@

With these changes we were able to generate grid-like patterns with
appropriate (linearly spaced) choices of the interaction parameters,
$\gamma$. To determine the $\gamma$ values suitable to recreate the cortical
barrel field by a principled method, we postulated that the interaction of
each thalamocortical axon growth cone with the ephrin gradients in the
cortical sub-plate might be determined by the location of the \emph{barreloid}
containing the axon's soma within signalling gradents \emph{in the
  thalamus}.
%That is, that interaction is grouped by barreloid identity.
%
%We therefore sought a map of the thalamic barreloids from the
%literature.
Ref.~\cite{haidarliu_size_2001} provided a comprehensive map of the
layout of the barreloids in rat, which is similar to mouse
\cite{van_der_loos_barreloids_1976}. We used these data to create the map of
52 barreloids shown in Fig.~\ref{fig}A. We postulated two linearly
varying molecules and used the barreloids' positions within this field to
generate the interaction parameters represented by the blue and red arrows.

We configured a simulation based on Eqs.~\ref{eq:dc}--\ref{eq:norm} to run on
a domain whose shape was matched to data from \cite{shimogori_fibroblast_2005}
(Fig.~\ref{fig}C). 52 separate axon branching fields were initialised with
random noise and given the $\gamma$ parameters from Fig.~\ref{fig}A. Two
orthogonal gradients were specified; molecule 1 with a posterior to anterior
gradient and molecule 2 with a medial-lateral gradient (resembling ephrin
A5). Fig.~\ref{fig}D--E shows snapshots of the connection density at
simulation times 0.08, 0.2 and 0.32 (${\delta}t=10^{-4}$), with colour
indicating the identity of the barreloid whose axons make maximum connections
at each location and \cmnt{contours showing where connection density passes
  0.5, indicating that each axon group is well localized}. A pattern emerges
which strongly resembles the barrel cortex in Fig.~\ref{fig}C, with `simulated
barrels' forming for whisker rows A to E, along with a suitably located group
for the rhinal whiskers.
%
Fig.~\ref{fig}G shows (using the Honda measure) that the simulated mechanism
does generate Voronoi patterns, in agreement with
\cite{senft_mouse_1991}.

% On stronger branching; not used in the results presented here.
% \cmnt{The realistic barrel pattern experiment is enhanced by stronger
%  branching competition:}
%
% To increase the strength of the competition, we choose an alternative form for
% $f(a)$ in Eq.~\ref{eq:comp} which has a geometric form for small $a$: $f(a) =
% \frac{1}{1 + e^{-m[a]^l}} - \frac{1}{2}$. The parameters $l$ and $m$ control
% the steepness and shape of the function.
%%% End of stronger branching competition %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A plausible theory of barrel patterning should reproduce the results of
experimental whisker barrel manipulations.
%
Here, we show that our mechanism reproduces the mirrored whisker barrel fields
reported in \cite{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012}
when an additional rostro-caudal source of Fgf8 is caused by
electroporation. We assume that a mechanism separate to ours governs the
definition of the boundary within which the barrel field will form and that
the additional Fgf8 source causes the boundary to become mirrored, with an
additional, posterior region forming. We also assume that under this
manipulation, the anteror-posterior guidance molecule (1) takes on a 'v'
shaped profile, with high expression at the far anterior and posterior of the
double barrel field domain, and low expression around the centre
(Fig.~\ref{fig}H). We assume molecule 2 is unchanged in its expression
pattern. Fig.~\ref{fig}I shows the pattern which forms with the simulation of
52 TC types having the same $\gamma$ parameters as set for the original
simulation (i.e.~from Fig.~\ref{fig}A). Similar, mirrored patterns form in
each side of the expanded domain. Note that for many of the 52 axon types, 2
fields form in spatially distinct locations (selected examples are shown),
though the straggler barrels $\alpha$ to $\delta$ and a few others (incl.~A1,
C1) appear in only single locations.

@@

We set out to test the hypothesis that barrel patterning may emerge via
self-organising processes.
%
We first extended an existing model of neocortical arealization to simulate
pattern formation in two dimensions, inheriting from this model the assumption
that axonal branching and synaptogenesis are coupled, and that thalamic
projections compete for cortical territory under the influence of molecular
signalling gradients.
%
Realistic barrel patterning emerged only when terms were added to represent
the assumption that thalamocortical projections compete for a limited supply
of connections with cortical cells, that their branching rates are reduced by
the branching of other thalamocortical projections, and that the density of
axon branching is conserved for each projection.
%
The net effect of these local interactions is a Voronoi tesselation in which
localised clusters of thalamic projections become somatotopically organised on
the cortical sheet according to the relative values of coefficients specifying
the rates at which axons are attracted or repelled by orthogonal molecular
signal gradients.
%
According to this model, the pattern of barrels is determined by the
competitive and cooperative interactions between cells relative to the shape
of the cortical field boundary. By computing this simulation on a realistic
barrel field, we provide an existence proof that the somatotopic order of the
barrels can be transferred without the need for absolute positional
information on their centre-point locations. Instead, relative interaction
information, extracted from the source tissue of the afferent axons (the
thalamus) can reproduce the somatotopy regardless of changes to the boundary
shape or size. A similar process could enable the transfer of peripheral
pattern information (the ultimate source of the pattern?) to the barrelettes
in the brainstem and from the barrelettes to the ventro basal thalamus.

The requirement for two orthogonal guidance gradients is easy to understand;
this provides a coordinate system to differentiate locations within the
cortical sub-plate. It is interesting to note that in order to successfully
reproduce the barrel patterns of Fig.~\ref{fig}C, we found it necessary for
axon types to experience \emph{both} positive \emph{and} negative interactions
with the two gradients. If only positive interactions are permitted (that is
growth cones only move up the gradient) then the pattern is heavily skewed
towards one side of the domain.
%
If empirical data showed that only positive (or negative) interactions were
possible, then by having \emph{four} guidance gradients, as two, orthogonal,
opposing pairs, the same, balancing effect could be achieved. Such a pattern
of opposing pairs of expression gradients is known to exist in the cortical
transcription factors that are related to Fgf8.

% Can of worms?
We think of this process as taking place in the cortical sub-plate and have
reported that the pattern resembles a Voronoi diagram. Anatomical data
suggests that the somatotopy is determined within the sub-plate. However, the
actual barrel pattern forms in layer IV of the cortex, to which axons are seen
to rise, and so the final Voronoi layout of the barrels is likely to be
influenced by mechanisms taking place within layer IV. This may explain why,
as the Voronoi measure of the simulated barrel pattern improves, the
correspondence between the simulated, and measured barrel layout diverges.

% Anything to say about the misexpression experiment?

\subsection*{Supplementary Material}

The code and configuration required to reproduce the results in this paper are
available at \url{https://github.com/ABRG-Models/BarrelEmerge}.
